---
title: 'Peer-graded Assignment: Course Project 1'
author: "Gabriel Paredes"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

### 1. Code for reading in the dataset and/or processing the data

Loading libraries for the analisys

```{r libraryLoad, echo=TRUE}
library(ggplot2)
library(lubridate)
library(dplyr)
```

Reading activity data and transforming colum date to date format
```{r readDat, aecho=TRUE}
unzip("activity.zip")
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
activity$date <- ymd(activity$date)
```

### 2. Histogram of the total number of steps taken each day

Constructs the Histogram
```{r plotHistogram, fig.align='center', echo=TRUE}
totalSteps <- activity %>% group_by(date) %>% summarise(Steps = sum(steps))

qplot(x = Steps, 
      data = totalSteps,
      geom = "histogram",
      bins = 5,
      col = I("black"), 
      alpha = 0.8,
      ylab = "Frequency", 
      xlab = "Steps/Day", 
      main = "Histogram total number of steps taken per day (5 bins)")
```

### 3.Calculate and report the mean and median of the total number of steps taken per day

```{r meanTotalSteps, echo=TRUE}
mean(totalSteps$Steps, na.rm = TRUE)
```

```{r medianTotalSteps, echo=TRUE}
median(totalSteps$Steps, na.rm = TRUE)
```

### 4. Time series plot of the average number of steps taken

```{r meanIntervalSteps, fig.align='center', echo=TRUE}
averageIntervalStep <- activity %>% group_by(interval) %>% summarise(Steps = mean(steps, na.rm = TRUE))

qplot(x = interval,
      y = Steps,
      data = averageIntervalStep,
      geom = "line",
      ylab = "Steps",
      xlab = "5 Minute Interval",
      main = "Average number of steps taken across all days"
      )
```

### 5. The 5-minute interval that, on average, contains the maximum number of steps

```{r maxInterval, echo=TRUE}
maxIntervalIndex <- which.max(averageIntervalStep$Steps)
maxInterval <- averageIntervalStep[maxIntervalIndex, ]
print(maxInterval)
```

The **`r maxInterval[[1, 1]]`** 5-minute interval, on average across all the days in the dataset, contains **`r maxInterval[[1, 2]]`** steps

### 6. Code to describe and show a strategy for imputing missing data

Total number of missing values in the dataset
```{r totalMissing, echo=TRUE}
missingIndex <- which(is.na(activity$steps))
length(missingIndex)
```

The missing values in the dataset are goin to be replaced with the mean for that 5-minute interval

```{r replaceNA, echo=TRUE}
fillActivity <- activity

for (i in missingIndex) {
    if (is.na(fillActivity$steps[i])) {
        imputedValue <- averageIntervalStep %>% filter(interval == fillActivity$interval[i]) %>% select(Steps)
        fillActivity$steps[i] <- imputedValue[[1]]
    }
}
```

### 7. Histogram of the total number of steps taken each day after missing values are imputed

Constructs the Histogram with the imputed values
```{r plotHistogramFill, fig.align='center', echo=TRUE}
totalStepsFill <- fillActivity %>% group_by(date) %>% summarise(Steps = sum(steps))

qplot(x = Steps, 
      data = totalStepsFill,
      geom = "histogram",
      bins = 5,
      col = I("black"), 
      alpha = 0.8,
      ylab = "Frequency", 
      xlab = "Steps/Day", 
      main = "Histogram total number of steps taken per day (5 bins) with imputed values")
```

Mean Total Steps per Day with imputed values
```{r meanTotalStepsFill, echo=TRUE}
mean(totalStepsFill$Steps)
```

Median Total Steps per Day with imputed values
```{r medianTotalStepsFill, echo=TRUE}
median(totalStepsFill$Steps)
```

The **mean** and **median** for the imputed values differs slightly from the original data set with the excluded NA values

### 7. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

Validates if the day is a weekday or a weekend
```{r weekend}
fillActivityDayType <- mutate(fillActivity, DayType = as.factor(ifelse(wday(date, label = TRUE) == "Sat" | wday(date, label = TRUE) == "Sun","Weekend", "Weekday")))
```

Panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r pannelPlot, fig.align='center', echo=TRUE}
averageIntervalStepDay <- fillActivityDayType %>% group_by(interval, DayType) %>% summarise(Steps = mean(steps))

qplot(x = interval,
      y = Steps,
      data = averageIntervalStepDay,
      geom = "line",
      ylab = "Steps",
      xlab = "5 Minute Interval",
      main = "Average number of steps taken across all days by Day Type",
      facets =  DayType ~ .,
      color = DayType
      )
```

